---
title: "SparXCC and clustering bi-partite networks"
output: html_notebook
---

This notebook describes the step to perform the SparXCC cross-correlation analysis between microbiome and host transcriptome.

The clustering of the bi-partite networks is also done with the Beckett method.

```{r}
#library(devtools)
library(CompoCor)
```

SparXCC uses counts data after filtering out low abundance OTUs, which are the same matrices used in reconstruction of co-occurrence networks.

Importing the OTU counts matrices:

```{r}
# Day samples
wallace_otu_counts_filtered_day <- read.csv("/home/renato/projects/fapesp_bepe_pd/microbiome/filtered_otu_table_day_filtered_rel_abund_cv_filtered.tsv", sep="\t", header=TRUE, row.names=1)
dim(wallace_otu_counts_filtered_day)
head(wallace_otu_counts_filtered_day)
colnames(wallace_otu_counts_filtered_day) <- gsub("^X", "", colnames(wallace_otu_counts_filtered_day))

# Night samples
wallace_otu_counts_filtered_night <- read.csv("/home/renato/projects/fapesp_bepe_pd/microbiome/filtered_otu_table_night_filtered_rel_abund_cv_filtered.tsv", sep="\t", header=TRUE, row.names=1)
dim(wallace_otu_counts_filtered_night)
head(wallace_otu_counts_filtered_night)
colnames(wallace_otu_counts_filtered_night) <- gsub("^X", "", colnames(wallace_otu_counts_filtered_night))
```


```{r}
kremling_expression_counts_filtered <- read.csv("/home/renato/projects/fapesp_bepe_pd/transcriptome/kremling_expression_v5_counts_filtered_cv_filtered.tsv", sep="\t", header=TRUE, row.names=1)
head(kremling_expression_counts_filtered)

kremling_expression_counts_filtered_day <- kremling_expression_counts_filtered[, grep("day", names(kremling_expression_counts_filtered), value = TRUE)]
kremling_expression_counts_filtered_night <- kremling_expression_counts_filtered[, grep("night", names(kremling_expression_counts_filtered), value = TRUE)]

head(kremling_expression_counts_filtered_day)
dim(kremling_expression_counts_filtered_day)
colnames(kremling_expression_counts_filtered_day) <- gsub("exp_day_", "", colnames(kremling_expression_counts_filtered_day))
head(kremling_expression_counts_filtered_day)

head(kremling_expression_counts_filtered_night)
dim(kremling_expression_counts_filtered_night)
colnames(kremling_expression_counts_filtered_night) <- gsub("exp_night_", "", colnames(kremling_expression_counts_filtered_night))
head(kremling_expression_counts_filtered_night)

wallace_otu_counts_filtered_day_sorted <- wallace_otu_counts_filtered_day[ , colnames(kremling_expression_counts_filtered_day)]
wallace_otu_counts_filtered_night_sorted <- wallace_otu_counts_filtered_night[ , colnames(kremling_expression_counts_filtered_night)]
```

Transposing dataframes:

```{r}
wallace_otu_counts_filtered_day_transposed <- t(wallace_otu_counts_filtered_day_sorted)
wallace_otu_counts_filtered_night_transposed <- t(wallace_otu_counts_filtered_night_sorted)
kremling_expression_counts_filtered_day_transposed <- t(kremling_expression_counts_filtered_day)
kremling_expression_counts_filtered_night_transposed <- t(kremling_expression_counts_filtered_night)
```

Making sure row names (samples) are identical between OTU and Expression data:

```{r}
identical(row.names(wallace_otu_counts_filtered_day_transposed), row.names(kremling_expression_counts_filtered_day_transposed))
identical(row.names(wallace_otu_counts_filtered_night_transposed), row.names(kremling_expression_counts_filtered_night_transposed))
```

Running SparXCC for the day and night samples:

```{r}
SparXCC_output_day <- SparXCC_base(wallace_otu_counts_filtered_day_transposed,
                        kremling_expression_counts_filtered_day_transposed,
                        pseudo_count = 1,
                        var_min = 1e-05,
                        Find_m = TRUE,
                        B_m = 100,
			            cores = 8)
write.table(SparXCC_output_day, file = "SparXCC_output_day.txt", sep = "\t")

SparXCC_output_night <- SparXCC_base(wallace_otu_counts_filtered_night_transposed,
                        kremling_expression_counts_filtered_night_transposed,
                        pseudo_count = 1,
                        var_min = 1e-05,
                        Find_m = TRUE,
                        B_m = 100,
			            cores = 8)
write.table(SparXCC_output_night, file = "SparXCC_output_night.txt", sep = "\t")
```

Since we noticed that the difference of numbers of cross-correlations between day and night samples is too high, we decided to run the SparXCC with the same plots for both day and night.

```{r}
# Find intersection of OTUs between day and night samples
common_plots <- intersect(colnames(wallace_otu_counts_filtered_day), colnames(wallace_otu_counts_filtered_night))
common_plots

# Select columns based on vector
wallace_otu_counts_filtered_day_common_plots <- wallace_otu_counts_filtered_day[ , common_plots]
wallace_otu_counts_filtered_night_common_plots <- wallace_otu_counts_filtered_night[ , common_plots]
kremling_expression_counts_filtered_day_common_plots <- kremling_expression_counts_filtered_day[ , common_plots]
kremling_expression_counts_filtered_night_common_plots <- kremling_expression_counts_filtered_night[ , common_plots]

head(wallace_otu_counts_filtered_day_common_plots)
head(kremling_expression_counts_filtered_day_common_plots)

# Check if columns are the same and in the same order
identical(colnames(wallace_otu_counts_filtered_day_common_plots), colnames(kremling_expression_counts_filtered_day_common_plots))
identical(colnames(wallace_otu_counts_filtered_night_common_plots), colnames(kremling_expression_counts_filtered_night_common_plots))

# Transpose dataframes
wallace_otu_counts_filtered_day_common_plots_transposed <- t(wallace_otu_counts_filtered_day_common_plots)
wallace_otu_counts_filtered_night_common_plots_transposed <- t(wallace_otu_counts_filtered_night_common_plots)
kremling_expression_counts_filtered_day_common_plots_transposed <- t(kremling_expression_counts_filtered_day_common_plots)
kremling_expression_counts_filtered_night_common_plots_transposed <- t(kremling_expression_counts_filtered_night_common_plots)
```

Running SparXCC with the same plots for both day and night samples:

```{r}
SparXCC_output_common_plots_day <- SparXCC_base(wallace_otu_counts_filtered_day_common_plots_transposed,
                        kremling_expression_counts_filtered_day_common_plots_transposed,
                        pseudo_count = 1,
                        var_min = 1e-05,
                        Find_m = TRUE,
                        B_m = 100,
			            cores = 8)
write.table(SparXCC_output_common_plots_day, file = "SparXCC_output_day_common_samples.txt", sep = "\t")

SparXCC_output_common_plots_night <- SparXCC_base(wallace_otu_counts_filtered_night_common_plots_transposed,
                        kremling_expression_counts_filtered_night_common_plots_transposed,
                        pseudo_count = 1,
                        var_min = 1e-05,
                        Find_m = TRUE,
                        B_m = 100,
			            cores = 8)
write.table(SparXCC_output_common_plots_night, file = "SparXCC_output_night_common_samples.txt", sep = "\t")
```

## Visualizing SparXCC results

Importing the edge list files for the day and night samples using the same/common plots (n = 174):

```{r}
sparxcc_day_common_samples <- read.csv("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_paper/SparXCC_output_day_common_samples_edgelist.tsv", sep="\t", header=TRUE)
#sparxcc_day_common_samples <- read.csv("/home/renato/projects/fapesp_bepe_pd/integration_transcriptome_microbiome/SparXCC_output_day_common_samples_edgelist.tsv", sep="\t", header=TRUE)
sparxcc_night_common_samples <- read.csv("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_paper/SparXCC_output_night_common_samples_edgelist.tsv", sep="\t", header=TRUE)
#sparxcc_night_common_samples <- read.csv("/home/renato/projects/fapesp_bepe_pd/integration_transcriptome_microbiome/SparXCC_output_night_common_samples_edgelist.tsv", sep="\t", header=TRUE)
```

Distribution of correlations from the edge list:

```{r}
library(ggplot2)

# Add a column to indicate positive or negative correlation
sparxcc_day_common_samples$correlation_type <- ifelse(sparxcc_day_common_samples$CorrelationCoefficient > 0, "Positive", "Negative")
sparxcc_night_common_samples$correlation_type <- ifelse(sparxcc_night_common_samples$CorrelationCoefficient > 0, "Positive", "Negative")

# Increase font size for axis labels, names, and title
theme_update(
  axis.title = element_text(size = 16),
  axis.text = element_text(size = 14),
  plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
)


# Plot histogram
png(filename = "day_sparxcc_dist.png")
ggplot(sparxcc_day_common_samples, aes(x = CorrelationCoefficient, fill = correlation_type)) +
    geom_histogram(binwidth = 0.01, color = "black", alpha = 0.7) +
    scale_fill_manual(values = c("Positive" = "blue", "Negative" = "red")) +
    labs(title = "Day",
             x = "Correlation (SparXCC)",
             y = "Frequency",
             fill = "Correlation Sign")
dev.off()

png(filename = "night_sparxcc_dist.png")
ggplot(sparxcc_night_common_samples, aes(x = CorrelationCoefficient, fill = correlation_type)) +
    geom_histogram(binwidth = 0.01, color = "black", alpha = 0.7) +
    scale_fill_manual(values = c("Positive" = "blue", "Negative" = "red")) +
    labs(title = "Night",
             x = "Correlation (SparXCC)",
             y = "Frequency",
             fill = "Correlation Sign")
dev.off()
```

