---
title: "Differential Co-expression analysis"
output: html_notebook
---

In addition to the network analysis carried out with corALS, a differential co-expression analysis was also carried out using csdR ([Pettersen & Almaas, 2022](https://link.springer.com/article/10.1186/s12859-022-04605-1)).

```{r}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("csdR")

#install.packages("igraph")
#install.packages("dplyr")

library(csdR)
library(magrittr)
library(igraph)
library(glue)
library(dplyr)
```

## Importing matrices

Matrices are TPM-normalized gene expression.

Before generating co-expression networks or compute the differential co-expression with the csdR package, we filtered the genes that are lowly expressed and/or have low variance (same genes used in co-expression and cross-correlation networks). 


```{r}
expression_kremling_tpm_filtered_night <- read.table("/home/renato/projects/fapesp_bepe_pd/transcriptome/expression_kremling_tpm_filtered_night.tsv", header = TRUE, sep = "\t", row.names = 1)
expression_kremling_tpm_filtered_day <- read.table("/home/renato/projects/fapesp_bepe_pd/transcriptome/expression_kremling_tpm_filtered_day.tsv", header = TRUE, sep = "\t", row.names = 1)
```

## Running csdR

Both matrices were transposed and used as input for the csdR package.
The function `run_csd` was used to compute the differential co-expression between day and night samples. The function was run with 1,000 iterations and 4 threads.

Note that the csdR package uses Spearman correlation coefficients.

```{r}
expression_kremling_tpm_filtered_day_transpose <- t(expression_kremling_tpm_filtered_day)
expression_kremling_tpm_filtered_night_transpose <- t(expression_kremling_tpm_filtered_night)

csd_results <- run_csd(
    x_1 = expression_kremling_tpm_filtered_day_transpose, x_2 = expression_kremling_tpm_filtered_night_transpose,
    n_it = 1000, nThreads = 4L, verbose = TRUE
)

save.image(file = ".diff_coexp_analysis_RData")
# Export to disk only if it is really necessary. R may die if the object is too big.
#write.csv(csd_results, "csd_results.csv", row.names = FALSE)
```


The interpretation of results for the differential co-expression analyses were carried out for the top 3000 pairs with the highest c, s, and d values.

Obtaining the top 3000 pairs for each c, s, and d values:

```{r}
pairs_to_pick <- 3000L
c_filter <- partial_argsort(csd_results$cVal, pairs_to_pick)
c_frame <- csd_results[c_filter, ]
s_filter <- partial_argsort(csd_results$sVal, pairs_to_pick)
s_frame <- csd_results[s_filter, ]
d_filter <- partial_argsort(csd_results$dVal, pairs_to_pick)
d_frame <- csd_results[d_filter, ]

csd_filter <- c_filter %>%
    union(s_filter) %>%
    union(d_filter)
csd_frame <- csd_results[csd_filter, ]

dim(csd_frame)
head(csd_frame)

write.csv(csd_frame, "csd_frame_top3000L.csv", row.names = FALSE)
```


Plotting the network using different colors for each type of edge (C, S, D):

```{r}
c_network <- graph_from_data_frame(c_frame, directed = FALSE)
s_network <- graph_from_data_frame(s_frame, directed = FALSE)
d_network <- graph_from_data_frame(d_frame, directed = FALSE)
E(c_network)$edge_type <- "C"
E(s_network)$edge_type <- "S"
E(d_network)$edge_type <- "D"
combined_network <- igraph::union(c_network, s_network, d_network)

join_attributes <- function(graph, attribute) {
    ifelse(
        test = is.na(edge_attr(graph, glue("{attribute}_1"))),
        yes = ifelse(
            test = is.na(edge_attr(graph, glue("{attribute}_2"))),
            yes = edge_attr(graph, glue("{attribute}_3")),
            no = edge_attr(graph, glue("{attribute}_2"))
        ),
        no = edge_attr(graph, glue("{attribute}_1"))
    )
}

E(combined_network)$edge_type <- join_attributes(combined_network, "edge_type")
layout <- layout_nicely(combined_network)
E(combined_network)$color <- recode(E(combined_network)$edge_type,
    C = "darkblue", S = "green", D = "darkred"
)
png("network_plot.png", width = 10000, height = 10000, res = 300)
plot(combined_network, layout = layout,
    vertex.size = 1, edge.width = 1, vertex.label.cex = 0.001, rescale = TRUE,
    width = 1000, height = 1000
)
while (!is.null(dev.list())) {
    dev.off()
}

```