---
title: "Co-occurrence networks for day and night samples in Wallace et al (2018) with paired RNA-Seq"
output: html_notebook
---

This notebook describes the reconstruction and analysis of co-occurrence networks of day and night samples in [Wallace et al (2018)](https://apsjournals.apsnet.org/doi/10.1094/PBIOMES-02-18-0008-R) that have paired RNA-Seq data.
OTUs are the same used in cross-correlations carried out between OTUs and maize genes.

Before reconstruction of co-occurrence networks, OTUs were filtered based on relative abundance (> 0.001 in at least 50% samples) and the coefficient of variation (OTUs above the first lower quartile).
The OTU counts are used in the construction of the networks.

```{r}
#library(devtools)
#install_github("zdk123/SpiecEasi")
library(SpiecEasi)
library(igraph)
```

```{r}
#Importing day and night tables
day_otu_filtered_sparxcc_wallace2018 <- read.csv('/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_paper/filtered_otu_table_day_filtered_rel_abund_cv_filtered.tsv', sep='\t', header=TRUE, row.names=1)
#day_otu_filtered_sparxcc_wallace2018 <- read.csv('/home/renato/projects/fapesp_bepe_pd/microbiome/filtered_otu_table_day_filtered_rel_abund_cv_filtered.tsv', sep='\t', header=TRUE, row.names=1)
colnames(day_otu_filtered_sparxcc_wallace2018) <- gsub("^X", "", colnames(day_otu_filtered_sparxcc_wallace2018))
night_otu_filtered_sparxcc_wallace2018 <- read.csv('/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_paper/filtered_otu_table_night_filtered_rel_abund_cv_filtered.tsv', sep='\t', header=TRUE, row.names=1)
#night_otu_filtered_sparxcc_wallace2018 <- read.csv('/home/renato/projects/fapesp_bepe_pd/microbiome/filtered_otu_table_night_filtered_rel_abund_cv_filtered.tsv', sep='\t', header=TRUE, row.names=1)
colnames(night_otu_filtered_sparxcc_wallace2018) <- gsub("^X", "", colnames(night_otu_filtered_sparxcc_wallace2018))

head(day_otu_filtered_sparxcc_wallace2018)
head(night_otu_filtered_sparxcc_wallace2018)

#Transposing matrices
day_otu_filtered_sparxcc_wallace2018_transposed <- t(day_otu_filtered_sparxcc_wallace2018)
night_otu_filtered_sparxcc_wallace2018_transposed <- t(night_otu_filtered_sparxcc_wallace2018)

#Running spiec-easi with Meinshausen-Bühlman (MB)
RNGkind("L'Ecuyer-CMRG")
set.seed(1234)
se.mb.day_otu_filtered_sparxcc_wallace2018 <- spiec.easi(day_otu_filtered_sparxcc_wallace2018_transposed, method='mb', lambda.min.ratio=1e-2, nlambda=20, pulsar.params=list(rep.num=50, ncores=8))
RNGkind("L'Ecuyer-CMRG")
set.seed(1234)
se.mb.night_otu_filtered_sparxcc_wallace2018 <- spiec.easi(night_otu_filtered_sparxcc_wallace2018_transposed, method='mb', lambda.min.ratio=1e-2, nlambda=20, pulsar.params=list(rep.num=50, ncores=8))

#Adding OTU names to adjacency matrices
se.mb.day_otu_filtered_sparxcc_wallace2018_matrix <- getRefit(se.mb.day_otu_filtered_sparxcc_wallace2018)
se.mb.night_otu_filtered_sparxcc_wallace2018_matrix <- getRefit(se.mb.night_otu_filtered_sparxcc_wallace2018)
rownames(se.mb.day_otu_filtered_sparxcc_wallace2018_matrix) <- colnames(day_otu_filtered_sparxcc_wallace2018_transposed)
rownames(se.mb.night_otu_filtered_sparxcc_wallace2018_matrix) <- colnames(night_otu_filtered_sparxcc_wallace2018_transposed)

# Extracting the adjacency matrix from spiec.easi object (day and night)
se.mb.day_otu_filtered_sparxcc_wallace2018.matrix <- symBeta(getOptBeta(se.mb.day_otu_filtered_sparxcc_wallace2018), mode='maxabs')
se.mb.day_otu_filtered_sparxcc_wallace2018.matrix.dsc <- se.mb.day_otu_filtered_sparxcc_wallace2018.matrix
se.mb.day_otu_filtered_sparxcc_wallace2018.matrix <- as.matrix(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix)
se.mb.night_otu_filtered_sparxcc_wallace2018.matrix <- symBeta(getOptBeta(se.mb.night_otu_filtered_sparxcc_wallace2018), mode='maxabs')
se.mb.night_otu_filtered_sparxcc_wallace2018.matrix.dsc <- se.mb.night_otu_filtered_sparxcc_wallace2018.matrix
se.mb.night_otu_filtered_sparxcc_wallace2018.matrix <- as.matrix(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix)

# Renaming rows and columns of the adjacency matrix (day and night)
rownames(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix) <- colnames(day_otu_filtered_sparxcc_wallace2018_transposed)
colnames(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix) <- colnames(day_otu_filtered_sparxcc_wallace2018_transposed)
day_sparxcc.otu.names <- colnames(day_otu_filtered_sparxcc_wallace2018_transposed)
rownames(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix) <- colnames(night_otu_filtered_sparxcc_wallace2018_transposed)
colnames(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix) <- colnames(night_otu_filtered_sparxcc_wallace2018_transposed)
night_sparxcc.otu.names <- colnames(night_otu_filtered_sparxcc_wallace2018_transposed)

# Generating the graph from the adjacency matrix (day and night)
se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net <- graph_from_adjacency_matrix(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix, mode = c("undirected"), weighted = TRUE, diag = FALSE)
V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$name <- day_sparxcc.otu.names
se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net <- graph_from_adjacency_matrix(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix, mode = c("undirected"), weighted = TRUE, diag = FALSE)
V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$name <- night_sparxcc.otu.names

# Converting edge weights into distances (larger values are shorter distances)
se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net.dist <- se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net
se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net.dist <- se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net
day_sparxcc.weights.dist <- 1 - abs(E(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net.dist)$weight)
night_sparxcc.weights.dist <- 1 - abs(E(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net.dist)$weight)
E(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net.dist)$weight <- day_sparxcc.weights.dist
E(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net.dist)$weight <- night_sparxcc.weights.dist
```

## Computing network metrics:

```{r}
# Extracting the number of edges from the networks
num_edges_day <- gsize(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)
num_edges_night <- gsize(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)

# Printing the number of edges
cat("Number of edges in the day network:", num_edges_day, "\n")
cat("Number of edges in the night network:", num_edges_night, "\n")

# Computing the average degree for the day and night networks
avg_degree_day <- mean(degree(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net))
avg_degree_night <- mean(degree(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net))

# Printing the average degree
cat("Average degree in the day network:", avg_degree_day, "\n")
cat("Average degree in the night network:", avg_degree_night, "\n")

# Computing the clustering coefficient for the day and night networks
clustering_coeff_day <- transitivity(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net, type = "average")
clustering_coeff_night <- transitivity(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net, type = "average")

# Printing the clustering coefficients
cat("Clustering coefficient in the day network:", clustering_coeff_day, "\n")
cat("Clustering coefficient in the night network:", clustering_coeff_night, "\n")

# Computing centralization closeness for the day and night networks
centralization_closeness_day <- centralization.closeness(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$centralization
centralization_closeness_night <- centralization.closeness(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$centralization

# Printing the centralization closeness
cat("Centralization closeness in the day network:", centralization_closeness_day, "\n")
cat("Centralization closeness in the night network:", centralization_closeness_night, "\n")

# Computing centralization betweenness for the day and night networks
centralization_betweenness_day <- centralization.betweenness(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$centralization
centralization_betweenness_night <- centralization.betweenness(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$centralization

# Printing the centralization betweenness
cat("Centralization betweenness in the day network:", centralization_betweenness_day, "\n")
cat("Centralization betweenness in the night network:", centralization_betweenness_night, "\n")
```

## Clustering the network (Louvain):

```{r}
RNGkind("L'Ecuyer-CMRG")
set.seed(1234)
day_sparxcc_wt <- cluster_louvain(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net, weights = E(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net.dist)$weight)
day_sparxcc_temp <- V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$name
day_sparxcc_temp <- as.data.frame(day_sparxcc_temp)
day_sparxcc_temp$louvain <- membership(day_sparxcc_wt)
day_sparxcc_wt

RNGkind("L'Ecuyer-CMRG")
set.seed(1234)
night_sparxcc_wt <- cluster_louvain(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net, weights = E(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net.dist)$weight)
night_sparxcc_temp <- V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$name
night_sparxcc_temp <- as.data.frame(night_sparxcc_temp)
night_sparxcc_temp$louvain <- membership(night_sparxcc_wt)
night_sparxcc_wt

V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain <- day_sparxcc_temp$louvain
V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)
V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain <- night_sparxcc_temp$louvain
V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)

# Plotting the day network with nodes colored by Louvain cluster
png(filename = "day_network_louvain_colored.1.png", width = 800, height = 800)
plot(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net, 
    vertex.color = V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain, 
    vertex.size = 5, 
    vertex.label = NA, 
    main = "Day Network: Louvain Clusters")
dev.off()

# Plotting the night network with nodes colored by Louvain cluster
png(filename = "night_network_louvain_colored.1.png", width = 800, height = 800)
plot(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net, 
    vertex.color = V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain, 
    vertex.size = 5, 
    vertex.label = NA, 
    main = "Night Network: Louvain Clusters")
dev.off()

# Just checking vertex attributes in the graph
vertex_attr(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)
vertex_attr(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)
```


## Computing Zi and Pi

This function requires a previous clustering step (above).

Using function from the [brainGraph R library](https://cran.r-project.org/web/packages/brainGraph/index.html):


```{r}
#install.packages("brainGraph")
library(brainGraph)
# Documentation: https://search.r-project.org/CRAN/refmans/brainGraph/html/vertex_roles.html
# Zi (within module connectivity)
Zi_day <- within_module_deg_z_score(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net, V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain)
Zi_night <- within_module_deg_z_score(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net, V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain)

V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$Zi <- Zi_day
V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$Zi <- Zi_night

# Pi (among-module connectivity)
Pi_day <- part_coeff(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net, V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain)
V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$Pi <- Pi_day
Pi_night <- part_coeff(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net, V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain)
V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$Pi <- Pi_night

se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net_df <- data.frame(Node = V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$name,
           Louvain_Module = V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain,
           Zi = V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$Zi,
           Pi = V(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net)$Pi)
se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net_df <- data.frame(Node = V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$name,
           Louvain_Module = V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$louvain,
           Zi = V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$Zi,
           Pi = V(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net)$Pi)

# Write dataframes to tabular files
write.table(se.mb.day_otu_filtered_sparxcc_wallace2018.matrix_net_df, file = "day_network_zi_pi_metrics.1.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
write.table(se.mb.night_otu_filtered_sparxcc_wallace2018.matrix_net_df, file = "night_network_zi_pi_metrics.1.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

# Adding titles to the plots
png(filename = "day_Pi_Zi_plot_with_title.1.png")
plot(Pi_day, Zi_day, xlab = "Among-module connectivity (Pi)", ylab = "Within-module degree z-score (Zi)", main = "Day Network: Pi vs Zi")
# Add threshold lines
abline(v = 0.62, col = "red", lty = 2, lwd = 2)  # Horizontal line at Pi = 0.62
abline(h = 2.5, col = "green", lty = 2, lwd = 2)  # Vertical line at Zi = 2.5
dev.off()

png(filename = "night_Pi_Zi_plot_with_title.1.png")
plot(Pi_night, Zi_night, xlab = "Among-module connectivity (Pi)", ylab = "Within-module degree z-score (Zi)", main = "Night Network: Pi vs Zi")
# Add threshold lines
abline(v = 0.62, col = "red", lty = 2, lwd = 2)  # Horizontal line at Pi = 0.62
abline(h = 2.5, col = "green", lty = 2, lwd = 2)  # Vertical line at Zi = 2.5
dev.off()
```

